---
title: 'LAB 3: Bias/Variance -- Exercise'
author: "Lluís A. Belanche, translated to R Markdown by Marta Arias"
date: "February 2020"
output:
  html_notebook: default
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---


## Exercise: Bias/Variance analysis on simulated data

```{r}
set.seed(5345)
par(mfrow=c(1, 1))
library(polynom)
```


Consider the (unknown) target function $f$ (and best solution to out problem):
```{r}
f <- function(x) sin(x-5)/(x-5)
```


From which we can sample datasets:

```{r}
N <- 150
x <- runif(N,0,15)           # generate the x according to a uniform distribution p(x)
t <- f(x) + rnorm(N, sd=0.1) # generate the t according to a gaussian conditional distribution p(t|x)
```


Plot the generated sample along with the (typically unknown) $f$
```{r}
plot(data.frame(x, t))
curve(f, type="l", col="blue", add=TRUE)
```

***

### Exercise for this lab session:


The exercise consists in estimating bias and variance (and hence bias^2+variance) for different models, and deduce which (polynomial) model is better for this problem.

To this end, you must generate many (thousands?) datasets of size $N$, choose one point $x \in [0,15]$ (I suggest $x=10$) and estimate bias and variance for it. Notice that you do not need to store the datasets. Remember that you can estimate these quantities using the Monte-Carlo method.

The models are going to be polynomials of degrees of your choice (I suggest 1,2,3,4,5,8).

