---
title: 'LAB 1: Data pre-processing exercise'
author: "Lluís A. Belanche, translated to R Markdown by Marta Arias"
date: "February 2020"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

<style>
p.comment {
background-color: #DBDBDB;
padding: 10px;
border: 1px solid black;
border-radius: 5px;
font-style: italic;
}
</style>


This exercise involves the use of the 'Auto' data set, which can be found in the file `Auto.data`. The file contains a number of variables for different cars.

In the following script you will find some steps that we have done for you, and at the end we list a series of tasks you should accomplish as practice for the pre-processing stage.


### Load data:

```{r}
graphics.off()      # reset/close all graphical devices 

# read the file 'Auto.data' (data on cars)
Auto <- read.table("Auto.data", header=TRUE, na.strings="?")
summary(Auto)
```

### Fix some stuff first:

```{r}
# put proper country of origin
Auto[,"origin"] <- factor(c("USA","EU","Japan")[Auto[,"origin"]])

# convert "miles per gallon" to "liters per km"
Auto[,"mpg"] <- 235.4/Auto[,"mpg"]
colnames(Auto)[which(colnames(Auto)=="mpg")] <- "l.100km"

summary(Auto)
```

### Car names:

The car name is not useful for modelling, but it may be handy to keep it as the row name

<p class="comment">
**WARNING!**  surprisingly, car names are not unique, so we first prefix them by their row number
</p>

```{r}
Auto$name <- paste (1:nrow(Auto), Auto$name)
rownames(Auto) <- Auto$name
Auto <- subset (Auto, select=-name)
head(Auto)
```

### Now we go for the cylinders

```{r}
table(Auto$cylinders)
```

That's strange, some cars have an odd number of cylinders (are these errors?)

```{r}
subset(Auto,cylinders==3)
```

These Mazdas wear a Wankel engine, so this is correct

```{r}
subset(Auto,cylinders==5)
```

Yes, these Audis displayed five-cylinder engines, so the data is correct

### Horsepower:
From `summary(Auto)` above we see that horsepower has 5 NA's that we'll need to take care of later ...

### So, this is your departing data set:

```{r}
summary(Auto)
```

### A plot from Lecture 1 ...

```{r}
with (Auto, Auto.lm <<- lm(l.100km ~ horsepower, Auto))

plot(Auto[,"horsepower"],Auto[,"l.100km"],
     pch=20,
     xlab="horsepower",ylab="fuel consumption (l/100km)",
     main="Linear regression")

# add regression line
a <- Auto.lm$coefficients["(Intercept)"]
b <- Auto.lm$coefficients["horsepower"]
abline(a=a,b=b,col="blue")
text(50,25,sprintf("y(x)=%.3fx+%.2f",b,a),col="red",pos=4)
```


### In order to create quick LaTeX code, try this:
  
```{r}
library(xtable)

xtable(Auto[1:4,])
xtable(Auto.lm)
```

Wasn't that nice? 

```{r}
# this is a list of R objects that can be embedded into a LaTeX table code:
methods(xtable)
```

***

### Exercise for the lab session:

1. Print the dimensions of the data set 

2. Identify possible target variables according to classification or regression problems

3. Inspect the first 4 examples and the predictive variables 6 and 7 for the tenth example

4. Perform a basic inspection of the dataset. Have a look at the minimum and maximum values for each variable; find possible errors and abnormal values (outliers); find possible missing values; decide which variables are really continuous and which are really categorical and convert them

5. Make a decision on a sensible treatment for the missing values and apply it; 

<p class="comment">
**WARNING!** `origin` is categorical and cannot be used for knn imputation, unless you make it binary temporarily
</p>

6. Derive one new continuous variable: weight/horsepower; derive one new categorical variable: sports_car, satisfying horsepower > 1.2*mean(horsepower) AND acceleration < median(acceleration); do you think this new variable is helpful in predicting 'origin' ?

7. Create a new dataframe that gathers everything and inspect it again

8. Perform a graphical summary of some of the variables (both categorical and continuous)

9. Perform a graphical comparison between some pairs of variables (both categorical and continuous)

10. Do any of the continuous variables "look" Gaussian? can you transform some variable so that it looks more so?

11. Create a new dataframe that gathers everything and inspect it again; consider 'origin' as the target variable; perform a basic statistical analysis as indicated in Section 9 of the example script

12. Shuffle the final dataset and save it into a file for future use

### Your code starts here ...

1. Print the dimensions of the data set.

```{r}
dim(Auto)
```

2. Identify possible target variables according to classification or regression problems.

```{r}
names(Auto)
```

We could target l.100km for regression, and year or origin for classification.

3. Inspect the first 4 examples and the predictive variables 6 and 7 for the tenth example.

```{r}
Auto[1:4,]
Auto[10,6:7]
```

4. Perform a basic inspection of the dataset. Have a look at the minimum and maximum values for each variable; find possible errors and abnormal values (outliers); find possible missing values; decide which variables are really continuous and which are really categorical and convert them.

```{r}
summary(Auto)
```
There are some outliers but they could be true values.
```{r}
Auto
```

