---
title: 'LAB 1: Data pre-processing'
author: "Lluís A. Belanche, translated to R Markdown by Marta Arias"
date: "February 2020"
output:
  html_notebook: default
  pdf_document: default
---

<style>
p.comment {
background-color: #DBDBDB;
padding: 10px;
border: 1px solid black;
border-radius: 5px;
font-style: italic;
}
</style>


Pre-processing is a necessary step prior to any analytical work that prepares and shapes a new dataset for further analysis. Each dataset is different and therefore it requires a different approach in what concerns data cleaning and preparation.
This pre-process is crucial because it can have a deep impact on future performance and results. Moreover, it can take a significant part of your time. As an informal guide, you shall pay attention to the following aspects (not necessarily in this order):

1. treatment of lost values (missing values)
2. treatment of anomalous values (outliers)
3. treatment of incoherent or incorrect values
4. elimination of irrelevant variables
5. (possible) elimination of redundant variables
6. coding of non-continuous or non-ordered variables (nominal or binary)
7. extraction of new variables that can be useful
8. normalization of the variables (e.g. standardization)
9. transformation of the variables (e.g. correction of skewness and/or kurtosis)

In this guide, we present examples of these pre-processing steps with the example dataset `credsco.csv` provided to you along with this notebook.


```{r}
# we will be using routines in the accompanying script:
source("auxiliary.R")
```

***

#### SECTION 1: Reading the data file `credsco.csv`

Reading properly a data set is non-trivial because you need to know how the data is formatted.
Things to look at are: what is the decimal separator, what is the column separator, is there a
header?, how are strings quoted, how (if any) are missing values encoded?, should character vectors be converted to factors?, should white spaces be stripped?, etc.

It is a good idea to consult the help on the reading command `?read.csv` and play with useful control parameters, like

```
# header=TRUE
# na.strings="?"
# dec = "."
# sep = ";"
# quote = "\"
```

After opening the file `credsco.csv` and inspecting it with any text editor, we decide on the following settings for the `read.csv` command:

```{r}
Credit = read.csv("credsco.csv", header = TRUE, quote = "\"", dec = ".", check.names=TRUE)
```

The dimensions of the data set are 4,455 examples described by 14 variables, which you can figure out with the `dim` command:

```{r}
dim(Credit)
```


This dataset contains data for clients of a bank, in particular regarding whether they have been assessed positively or negatively for a credit loan. This dataset is typically used to train classifiers that predict whether a client will be granted a loan or not (so, the target variable is `Assessment`; it has two possible values -- however we observe a strange row with a 3rd possible value).

```{r}
# what are the variables?
names(Credit)
```


```{r}
# count frequence of values for target variable
table(Credit$Assessment)
```



```{r}
# inspect the first 4 examples
Credit[1:4,]
```

```{r}
# inspect predictive variables 4, 5, 6 and 7 for the first example
Credit[1,5:8]
```

***

#### SECTION 2: Basic inspection of the dataset

Have a look at the minimum and maximum values for each variable; find possible errors and abnormal values (outliers); find possible missing values; decide which variables are continuous and which are categorical; if there are mixed types, we have three options: recode continuous to categorical, recode categorical to continuous or leave them as they are. Try to get a first idea of the best decision for each issue.

```{r}
# always have a look at basic statistics of each variable in your data
summary(Credit)
```

From the summary, we observe the following:

* `Assessment`,`Housing`,`MaritalStatus`,`Records`,`TypeOfJob` are categorical and need to be treated properly. In particular, `Assessment` is the target variable; we need to identify correct values

* `Capital`, `ChargesOnCapital` and `Income` present abnormally high maximums (99999999)

* there are also suspicious zeros, in both types of variables, which we identify with missing values

***

#### SECTION 3: Dealing with **_missing values_**

Make a decision on a sensible treatment for the missing values and apply it; it is wise to write down the possible consequences of this decision and the alternatives that could be considered in case the final results are not satisfactory

The easiest way is of course to eliminate the involved rows or columns; this can be done partially. For example, we could decide to eliminate the variables with the highest proportion of missing values. Deleting instances and/or variables containing missing values results in loss of relevant data and is also frustrating because of the effort in collecting the sacrificed information.

**CAREFUL!** R does not know magically which entries are missing values

```{r}
# this does not do anything, because there are no explicitly declared NA's
Credit.complete = na.omit(Credit)
dim(Credit.complete)
```

The previous code does nothing, because missing values are not encoded as NA in our data.

In the present case we have decided to perform a step-by-step treatment, separate for the categorical and continuous information.

First, we decide to remove those rows with with missing values in the categorical variables (there are few), which in our data are encoded with a `0`.

```{r}
table(Credit[,"Assessment"]==0)
table(Credit[,"Housing"]==0)
table(Credit[,"MaritalStatus"]==0)
table(Credit[,"TypeOfJob"]==0)
```

Since there aren't too many, we remove those rows with a 0 in these variables:
```{r}
Credit = Credit[Credit[,"Assessment"] != 0 & Credit[,"Housing"] != 0 & 
                 Credit[,"MaritalStatus"] != 0 & Credit[,"TypeOfJob"] != 0,]

dim(Credit)
```

This process removed `r 4455 - 4446` rows.

Now we process rows with missing values in the continuous variables (code 99999999). To avoid having to write each time the `Credit$...` part, we use the `attach` command. 

```{r}
attach(Credit)
```

However, beware:

<p class="comment">
**WARNING!**  `attach` allows access to the values of columns of a data frame for reading (access) only, and as they were when attached, but does not modify the dataframe
</p>

```{r}
hist(Income)
hist(Income[Income != 99999999])
hist(Income[Income != 99999999 & Income != 0], breaks=15)
```


These values are clearly incorrect!

```{r}
table(Income == 99999999)
table(Income == 0)
table(Capital == 99999999)
table(ChargesOnCapital == 99999999)
```

What do we do with this one? well, let's assume it is correct ...

```{r}
table(YearsInJob == 0)
```

Continuous variables have too many missing values, we can not eliminate them just like that: we must devise a treatment for these missing values (we will **impute** the missing values, see below).

First, we declare them as `NA`, including those from no `Income`

```{r}
Income[Income == 99999999 | Income == 0] = NA
Capital[Capital == 99999999] = NA
ChargesOnCapital[ChargesOnCapital == 99999999] = NA
```

```{r}
# see the difference
summary(Income)
summary(Credit$Income)
```

##### Imputation of NA's for continuous variables

Imputation by **nearest neighbour**: for every observation with a missing `Income`, we look for the most similar observation (according to the remaining variables) and copy its `Income` value.

```{r}
# Imputation of 'Income', 'Capital' and 'ChargesOnCapital'
Income = knn.imputation (Income, "Income")
Capital = knn.imputation (Capital, "Capital")
ChargesOnCapital = knn.imputation (ChargesOnCapital, "ChargesOnCapital")

ChargesOnCapital[Capital==0] = 0
```

<p class="comment">
Don't forget to assign back to the dataframe
</p>

```{r}
Credit[,"Income"] = Income
Credit[,"Capital"] = Capital
Credit[,"ChargesOnCapital"] = ChargesOnCapital
```

Inspect again the result, especially the new statistics

```{r}
dim(Credit)
summary(Credit)
```

***

#### SECTION 4: Treatment of mixed data types

First we explicitly declare some variables as categorical ('factor' in R)

```{r}
Assessment = as.factor(Assessment)
Housing = as.factor(Housing)
MaritalStatus = as.factor(MaritalStatus)
Records = as.factor(Records)
TypeOfJob = as.factor(TypeOfJob)

levels(Assessment)
levels(Housing)
levels(MaritalStatus)
levels(Records)
levels(TypeOfJob)
```

These level codes are not very nice (not very meaningful), let's recode:

```{r}
levels(Assessment) = c("positive","negative")
levels(Housing) = c("rent","owner","private","ignore","parents","other")
levels(MaritalStatus) = c("single","married","widower","split","divorced")
levels(Records) = c("no","yes")
levels(TypeOfJob) = c("indefinite","temporal","self-employed","other")
```


As a reminder of dataframe access (i.e., behaviour with `attach`):

```{r}
is.factor(Assessment)
is.factor(Credit[,"Assessment"])
```

***

#### SECTION 5: Derivation of new variables (a.k.a. feature extraction)

Decide whether it can be sensible to derive new variables; we decide to extract two new continuous and one new categorical variable (for the sake of illustration):

```{r}
# Financing ratio (continuous)
FinancingRatio = 100*AmountRequested/MarketPrice
hist(FinancingRatio)
```

```{r}
# Saving capacity (continuous)
SavingCapacity = (Income-Expenses-(ChargesOnCapital/100))/(AmountRequested/Deadline)
hist(SavingCapacity, breaks=16)
```


```{r}
# Amount Requested greater than 1.2 times the median by people 1.5 times older than the mean (categorical):
Dubious = rep("No", nrow(Credit))
Dubious[AmountRequested > 1.2*median(AmountRequested, na.rm = TRUE) & Age > 1.5*mean (Age, na.rm = TRUE)] = "Yes"
Dubious = as.factor(Dubious)

table(Dubious,Assessment)
```


***

#### SECTION 6: Putting together what we have done so far

```{r}
# create a new dataframe that gathers everything and inspect it again

Credit.new = data.frame(Assessment,YearsInJob,Housing,Deadline,Age,MaritalStatus,Records,TypeOfJob,Expenses,Income,Capital,ChargesOnCapital,AmountRequested,MarketPrice,FinancingRatio,SavingCapacity,Dubious)
                   
summary(Credit.new)
dim(Credit.new)

detach(Credit)
attach(Credit.new)
```

Sanity check:

```{r}
is.factor(Credit.new[,"Assessment"])
```

***

#### SECTION 7: Gaussianity and transformations

It is useful to have an idea of the empirical distribution of the dataset's variables. One can visualize a graphical summary of some of the variables (both categorical and continuous), using the `boxplot()` and `hist()` procedures.

##### For **continuous** data: histograms and boxplots


```{r}
hist(Income, col=2)
```

```{r}
par(mfrow=c(1,2))
hist(Capital)
hist(log10(Capital), breaks=20)
```

```{r}
boxplot(Deadline)
title ("These are the deadlines")
```

```{r}
boxplot(Age, col="lightgray")
title ("and these are the ages")
```

```{r}
boxplot(Credit.new[,9:16], outline=TRUE) 
boxplot(Credit.new[,9:16], outline=FALSE) # much better, but would be nicer one by one, to avoid flat boxplots
```

The previous plots suggest to take logs on some variables: `Capital` and `ChargesOnCapital` (we'll do it later)

##### For categorical data: Frequency tables, Contingency tables, Bar charts, Pie charts

Now, should we treat Age as categorical? probably not ... but if we wanted to do so, we can use `cut`:

```{r}
table(Age)                            
summary(Age)
(Age.cat = cut(Age, breaks = seq(30, 90, 10))) # WARNING! we are generating NAs
```

```{r}
Age.cat = cut(Age, breaks = seq(15, 75, 10))   
(Age.tab = table(Age.cat))
```


```{r}
barplot(Age.tab)    # bar chart
pie(Age.tab)        # pie chart
```

Incidentally, this is how we could generate another new variable based on Age:

```{r}
Age2.cat = factor(as.integer(Age < 55))        
levels(Age2.cat) = c("over55","under55")
```


With the `table` commands we can generate contingency tables for categorical variables in different flavors:

```{r}
(TypeOfJob.Age = table(TypeOfJob, Age.cat)) # contingency table
margin.table(TypeOfJob.Age, 1)              # row sums
margin.table(TypeOfJob.Age, 2)              # column sums

round(prop.table(TypeOfJob.Age), digits=3)  # relative freqencies, rounded to 3 digits 
round(prop.table(TypeOfJob.Age) * 100, digits=1)  # total percentages

round(prop.table(TypeOfJob.Age, 1), digits=3) # table of relative frequencies (row-wise)
(TypeOfJob.Age.rel = round(prop.table(TypeOfJob.Age, 2), digits=3)) # table of relative frequencies (column-wise)
```

Other graphical representations of the distributions:

```{r}
barplot(TypeOfJob.Age)  # basic stacked bar chart
```

```{r}
barplot(TypeOfJob.Age.rel, yaxt="n", xlab="Age", ylab="proportion", 
        col = c("white", "grey80", "grey40", "black"), 
        main = "TypeOfJob by Age", xlim=c(0,9))   # stacked bar chart

legend("bottomleft", legend=rownames(TypeOfJob.Age.rel), col="black", 
      fill = c("white", "grey80", "grey40", "black"), cex=0.65)
```

```{r}
barplot(TypeOfJob.Age.rel, beside = TRUE)   # grouped bar chart
axis(2, at=seq(0, 1, 0.2))
```

#### Pair-wise comparisons of variables

We can also perform graphical comparisons between some pairs of variables (both categorical and continuous), using the plot(),  pairs() and identify() procedures

```{r}
par(mfrow=c(1,2))
plot (AmountRequested, Capital, main = "Amount req. vs. Market price", 
      cex = .5, col = "dark red")

plot (log10(AmountRequested), log10(Capital+1), main = "Amount req. vs. Market price (better)", cex = .5, col = "dark red")

# adding a center (dashed) and a regression line (blue)
abline(v = mean(log10(AmountRequested)), lty = 2, col = "grey40")
abline(h = mean(log10(Capital+1)), lty = 2, col = "grey40")
abline(lm(log10(Capital+1) ~ log10(AmountRequested)), col = "blue")
```

Note that log10(x+1)=0 for x=0, so our transformation keeps the zeros. On the other hand, these same zeros spoil the regression

```{r}
par(mfrow=c(1,1))

plot (TypeOfJob, AmountRequested)

plot (Age,Assessment, col="red") # WARNING!
plot (Assessment, Age, col="red") # better

plot (Assessment, TypeOfJob,col="blue", xlab="Assessment",ylab="TypeOfJob")

pairs(~ AmountRequested + Capital + Age)
```

```{r}
# Plotting a variable against the normal pdf in red
hist.with.normal(Expenses, "Expenses")
```

Density plots for other variables:

```{r}
par(mfrow=c(2,4))
for (i in 0:7) 
  { plot(density(Credit.new[,i+9]), xlab="", main=names(Credit.new)[i+9]) }
```

Do any of the continuous variables "look" Gaussian? 
Features to look for in comparing to a Gaussian: outliers, asymmetries, long tails

A useful tool for "Gaussianization" is the Box-Cox power transformation

```{r}
library(MASS)

par(mfrow=c(1,3))

hist(Capital, main="Look at that ...")

bx = boxcox(I(Capital+1) ~ . - Assessment, data = Credit.new,
             lambda = seq(-0.25, 0.25, length = 10))

lambda = bx$x[which.max(bx$y)]

Capital.BC = (Capital^lambda - 1)/lambda

hist(Capital.BC, main="Look at that now!")

par (mfrow=c(1,1))
```

***

#### SECTION 8: Dealing with mixed types of variables (revisited)

##### Recoding categorical variables (dummy binary coding a,k.a. one-hot encoding)

Most R functions for modelling accept factors mixed with numeric data; the factors are recoded using a dummy binary code, so there is no need to handle this explicitly

##### Recoding continuous variables (a.k.a. discretization or categorization or factorization)

```{r}
YearsInJob.CAT         = cut(YearsInJob, breaks=c(-1,1,3,8,14,99))
Deadline.CAT           = cut(Deadline, breaks=c(0,12,24,36,48,99))
Age.CAT                = cut(Age, breaks=c(0,25,30,40,50,99))
Expenses.CAT           = cut(Expenses, breaks=c(0,40,50,60,80,9999))
Income.CAT             = cut(Income, breaks=c(0,80,110,140,190,9999))
Capital.CAT            = cut(Capital, breaks=c(-1,0,3000,5000,8000,999999))
ChargesOnCapital.CAT   = cut(ChargesOnCapital, breaks=c(-1,0,500,1500,2500,999999))
AmountRequested.CAT    = cut(AmountRequested, breaks=c(0,600,900,1100,1400,99999))
MarketPrice.CAT        = cut(MarketPrice, breaks=c(0,1000,1300,1500,1800,99999))
FinancingRatio.CAT     = cut(FinancingRatio, breaks=c(0,50,70,80,90,100))
SavingCapacity.CAT     = cut(SavingCapacity, breaks=c(-99,0,2,4,6,99))

levels(YearsInJob.CAT)         = paste("YearsInJob",levels(YearsInJob.CAT))
levels(Deadline.CAT)           = paste("Deadline",levels(Deadline.CAT))
levels(Age.CAT)                = paste("Age",levels(Age.CAT))
levels(Expenses.CAT)           = paste("Expenses",levels(Expenses.CAT))
levels(Income.CAT)             = paste("Income",levels(Income.CAT))
levels(Capital.CAT)            = paste("Capital",levels(Capital.CAT))
levels(ChargesOnCapital.CAT)   = paste("ChargesOnCapital",levels(ChargesOnCapital.CAT))
levels(AmountRequested.CAT)    = paste("AmountRequested",levels(AmountRequested.CAT))
levels(MarketPrice.CAT)        = paste("MarketPrice",levels(MarketPrice.CAT))
levels(FinancingRatio.CAT)     = paste("FinancingRatio",levels(FinancingRatio.CAT))
levels(SavingCapacity.CAT)     = paste("SavingCapacity",levels(SavingCapacity.CAT))
```

```{r}
Credit.new.CAT = data.frame(Assessment,YearsInJob.CAT,Housing,Deadline.CAT,Age.CAT,MaritalStatus,Records,TypeOfJob,Expenses.CAT,Income.CAT,Capital.CAT,ChargesOnCapital.CAT,AmountRequested.CAT,MarketPrice.CAT,FinancingRatio.CAT,SavingCapacity.CAT,Dubious)
```

Notice the difference:

```{r}
names(Credit.new)
Credit.new[1,]
dim(Credit.new)

names(Credit.new.CAT)
Credit.new.CAT[1,]
dim(Credit.new.CAT)
```

***

#### SECTION 9: Statistical analysis

```{r}
summary(Credit.new) # yet again
```

##### Basic summary statistics by groups

Plot variable statistics for each class using `describeBy`:

```{r}
library(psych) # describeBy

range.cont = c(2,4,5,9:16)

describeBy(Credit.new[,range.cont], Credit.new$Assessment)
```

Plot of all pairs of some continuous variables according to the class:

```{r}
# (this plot shows how difficult this problem is)
pairs(Credit.new[,c(2,5,9,16)], main = "Credit Scoring DataBase", col = (1:length(levels(Credit.new$Assessment)))[unclass(Credit.new$Assessment)])
```
                        
##### Feature selection for continuous variables using Fisher's F


```{r}
names(Credit.new)[range.cont]

pvalcon = NULL

varc = list(YearsInJob, Deadline, Age, Expenses, Income, Capital, ChargesOnCapital, AmountRequested, MarketPrice, FinancingRatio, SavingCapacity)

for (i in 1:length(varc)) 
  pvalcon[i] = (oneway.test(varc[[i]]~Credit.new$Assessment))$p.value

pvalcon = matrix(pvalcon)
row.names(pvalcon) = colnames(Credit.new)[range.cont]

# Ordered list of continuous variables according to their association to Assessment (more to less association)

as.matrix(sort(pvalcon[,1]))

# Graphical representation of Assessment

ncon = nrow(pvalcon)

par (mfrow=c(3,4))

for (i in 1:ncon) 
{
  barplot (tapply(varc[[i]], Credit.new$Assessment, mean),main=paste("Means by",row.names(pvalcon)[i]), las=2, cex.names=1.25)
  abline (h=mean(varc[[i]]))
  legend (0,mean(varc[[i]]),"Global mean",bty="n") 
}

```

##### Feature selection using pair-wise Pearson correlations

```{r}
# Compute and show the 10 most mosthighly correlated variables
mosthighlycorrelated(Credit.new[,range.cont], 10)
```

*** 

#### SECTION 9: Miscellaneous issues

Shuffle the data (to avoid possible biases)

```{r}
set.seed (104)
Credit.new = Credit.new[sample.int(nrow(Credit.new)),]
``` 

***

#### SECTION 10: Save the preprocessed data into a file for future use

```{r}
# WARNING! This is a binary file, unless 'ascii' is TRUE
save(Credit.new, file = "Credsco-processed.Rdata")
```

Cleanup:

```{r}
# remove (almost) everything in the working environment
rm(list = ls())

load("Credsco-processed.Rdata")
objects()
```

