---
title: 'LAB 7: GLMs - Logistic Regression and beyond'
author: "Lluís A. Belanche, translated to R Markdown by Marta Arias"
date: "April 2020"
output:
  html_notebook: default
pdf_document: default
editor_options: 
  chunk_output_type: inline
---

### Exercise 1

This first exercise involves the use of the South African Coronary Heart Disease (CHD) data set (the one mentioned in the lecture slides).

The task is to predict CHD using age at onset, current alcohol consumption, obesity levels, cumulative tobacco, type-A behavior, and low density lipoprotein cholesterol as predictors.

```{r}
SAheart <- read.csv("MDSAheart.data")
```

First some necessary pre-processing (this part should be lenghtier as you know but let's go to the point):
```{r}
SAheart <- subset(SAheart, select=-row.names)
SAheart[,"famhist"] <- as.factor(SAheart[,"famhist"])
SAheart[,"chd"] <- factor(SAheart[,"chd"], labels=c("No","Yes"))
```


There seem to be no missing values;
'alcohol', 'ldl' and 'tobacco' have very skewed distributions! a log transform could be in order ... it seems that quite a bit of people have large values (not surprising)

So this is your departing data set:

```{r}
summary(SAheart) 
```


Do the following (in separate sections):

1. Perform a pairs() plot of the variables paired for a first exploration, with chd overlayed (in color)
2. Perform various boxplots (notice that all but one of the predictors are continuous)
3. Decide if you log-transform some of the continuous variables
4. Create learning and test sets, selecting randomly 2/3 and 1/3 of the data
5. Fit a logistic regression model using only the learn data with Coronary Heart Disease (chd) as the target
  and observe the results (summary):
    - build a maximal model using the learn data
    - try to simplify the model by eliminating the least important variables progressively
    - refit the model using these variables
6. Interpret a couple of coefficients
7. Calculate the misclassification rate for your model in the learn data and in the leftout data (test data)
  Try to come up with a more acceptable model by playing with how the classifier assigns classes according to probability

```{r}
# Your code starts here ...

```

***

### Exercise 2

The following data represents the number of news AIDS cases in Belgium, from 1981 onwards.
Data from the book by Venables & Ripley (Modern applied statistics in S).

```{r}
Year <- 1:13
Cases <- c(12,14,33,59,67,74,123,141,165,204,253,246,240)

plot (Year+1980, Cases, col="blue", xlab="Year", ylab="News AIDS cases")
```


1. Justify the fitting of a Poisson regression model
2. Fit the Poisson regression model and examine the results
3. Give the model using the obtained coefficients
4. Gather predictor, outcome and prediction together and plot everything
5. Analyse the residuals and decide a derived predictor; redo the model and compare


```{r}
# Your code starts here ...

```


