---
title: 'LAB 7: GLMs - Logistic Regression and beyond'
subtitle: 'Starter: Maximum Likelihood'
author: "Lluís A. Belanche, translated to R Markdown by Marta Arias"
date: "April 2020"
output:
  html_notebook: default
pdf_document: default
editor_options: 
  chunk_output_type: inline
---


First we are going to play with a very simple example of Maximum Likelihood for the binomial distribution. The goal is to estimate the probability $p$ of getting heads in $N$ coin tosses if we get $n1$ heads (exactly as seen in class)

```{r}
set.seed(1234)
```

Suppose we toss a coin $N=100$ times and we get $n1=70$ heads

```{r}
N <- 100
n1 <- 70
```

We generate all possible values for $p$ and put them in a vector:
```{r}
p <- seq(from=1e-5, to=1, by=0.01)
```


The likelihood function (as a function of p) is:
```{r}
L <- choose(N,n1)*p^n1*(1-p)^(N-n1)
```

Next, let us visualize the likelihood fn:

```{r}
plot.L <- function (L, text)
{
  plot(p, L, type="l", ylab=text,xaxt = "n")
  grid(nx=10)
  axis(side = 1, at = seq(0,1,by=0.1), las = 2, hadj = 0.9)
  abline(v=0.7, col="red")
}

plot.L (L,"likelihood of p")
```

Remember that oftentimes instead of maximizing the likelihood function we maximize the *log likelihood* instead. Let us
visualize this as well:

```{r}
logL <- log(choose(N,n1)) + n1*log(p) + (N-n1)*log(1-p)
plot.L (logL, "log-likelihood of p")
```

Notice that it also tops at 0.7 (it has too since `log` is a strictly monotone function)
Note also that `log()` is the natural logarithm in `R`.


The maximum is attained at n1/N = 0.7 (the proportion of heads)