---
title: 'LAB 8: Multilayer Perceptrons'
subtitle: 'Example 2: circular artificial 2D data'
author: "Lluís A. Belanche, translated to R Markdown by Marta Arias"
date: "April 2020"
output:
  html_notebook: default
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

This script will use artifically generated data. It consists of two concentric classes and it is therefore not separable by a linear classifier. Here is the data we generate.

```{r}
library(MASS)
library(nnet)

set.seed(3)
par(mfrow=c(1,1))

p <- 2
N <- 200

x <- matrix(rnorm(N*p),ncol=p)
y <- as.numeric((x[,1]^2+x[,2]^2) > 1.4)
mydata <- data.frame(x=x,y=y)
plot(x, col=c('black','green')[y+1], pch=19, asp=1)
```


Let's start by using one hidden layer with 3 hidden units, no regularization and the error function "cross-entropy":

Note: In this case it is not necessary to standardize because they variables already are (they have been generated from a distribution with mean 0 and standard deviation 1).

```{r}
set.seed(3)
nn1 <- nnet(y~x.1+x.2, data=mydata, entropy=T, size=3, decay=0, maxit=2000, trace=F)

yhat <- as.numeric(predict(nn1,type='class'))
par(mfrow=c(1,2))
plot(x,pch=19,col=c('black','green')[y+1],main='actual labels',asp=1)
plot(x,col=c('black','green')[(yhat>0.5)+1],pch=19,main='predicted labels',asp=1)
table(actual=y,predicted=predict(nn1,type='class'))
```


Excellent, indeed!

Let's execute it again, this time wth a different random seed


```{r}
set.seed(9)

nn1 <- nnet(y~x.1+x.2, data=mydata, entropy=T, size=3, decay=0, maxit=2000, trace=F)
yhat <- as.numeric(predict(nn1,type='class'))
par(mfrow=c(1,2))
plot(x,pch=19,col=c('black','green')[y+1],main='actual labels',asp=1)
plot(x,col=c('black','green')[(yhat>0.5)+1],pch=19,main='predicted labels',asp=1)
table(actual=y,predicted=predict(nn1,type='class'))
```


We see that the optimizer does not always find a good solution

***

How many hidden units do we need?

```{r}
par(mfrow=c(2,2))
for (i in 1:4)
{
  set.seed(3)
  nn1 <- nnet(y~x.1+x.2, data=mydata, entropy=T, size=i, decay=0, maxit=2000, trace=F)
  yhat <- as.numeric(predict(nn1,type='class'))
  plot(x,pch=20,col=c('black','green')[yhat+1])
  title(main=paste('nnet with',i,'hidden unit(s)'))
}
```

***

Let's find out which function has been learned exactly, with 3 units:

```{r}
set.seed(3)
nn1 <- nnet(y~x.1+x.2, data=mydata, entropy=T, size=3, decay=0, maxit=2000, trace=F)

# create a grid of values
x1grid <- seq(-3,3,l=200)
x2grid <- seq(-3,3,l=220)
xg <- expand.grid(x1grid,x2grid)
xg <- as.matrix(cbind(1,xg))

## input them to the hidden units, and get their outputs
h1 <- xg%*%matrix(coef(nn1)[1:3],ncol=1)
h2 <- xg%*%matrix(coef(nn1)[4:6],ncol=1)
h3 <- xg%*%matrix(coef(nn1)[7:9],ncol=1)

## this is the logistic function, used by nnet() for the hidden neurons, and 
## for the output neurons in two-class classification problems
logistic <- function(x) {1/(1+exp(-x))}

## the hidden units compute the logistic function, so we cut the output value at 0.5; we get a decision line

par(mfrow=c(2,2))
contour(x1grid, x2grid, matrix(h1, 200, 220), levels=0.5)
contour(x1grid, x2grid, matrix(h2, 200, 220), levels=0.5, add=T)
contour(x1grid, x2grid, matrix(h3, 200, 220), levels=0.5, add=T)
title(main='net input = 0.5\n in the hidden units')

z <- coef(nn1)[10] + coef(nn1)[11]*logistic(h1) + coef(nn1)[12]*logistic(h2) + coef(nn1)[13]*logistic(h3)

contour(x1grid,x2grid,matrix(z,200,220))
title('hidden outputs = logistic of the net inputs\n and their weighted sum')
contour(x1grid,x2grid,matrix(logistic(z),200,220),levels=0.5)
title('logistic of the previous sum')
contour(x1grid,x2grid,matrix(logistic(z),200,220),levels=0.5)
points(x,pch=20,col=c('black','green')[y+1])
title('same with training data points')
```