---
title: 'LAB 8: Multilayer Perceptrons'
author: "Lluís A. Belanche, translated to R Markdown by Marta Arias"
date: "April 2020"
output:
  html_notebook: default
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

### Exercise 1

We continue with the mixture of distributions artificial 2D data from example script 3. In order to find the best network architecture, you are asked to explore the two methods:

a) Explore different numbers of hidden units in one hidden layer, with no regularization
b) Fix a large number of hidden units in one hidden layer, and explore different regularization values

Very much as it was done in the first script; you need first to split the data into training (2/3) and test (1/3)

Once you select the best model overall (including the two methods), compute the test error and plot the learned decision function against the data and the truth


### Exercise 2

This exercise is based on Doppler's function:

```{r}
doppler <- function (x) { sqrt(x*(1-x))*sin(2.1*pi/(x+0.05)) }
```


We are going to model this function in the interval (0.2,1)

```{r}
a <- 0.2
b <- 1

doppler.data <- function (N, a, b) 
{
  x <- runif(N, a, b)
  t <- doppler(x) + rnorm(N, sd=0.1)
  dd <- data.frame(x,t)
  names(dd) <- c("x", "t")
  return (dd)
}

d <- doppler.data (500, a , b)
```



The black points are the data, the blue line is the true underlying function

```{r}
par(mfrow=c(1,1))

plot (d)
curve (doppler(x), a, b, col='blue', add=TRUE)
```

Your tasks:

1. Set `N = 100` and model the data using `nnet()`; compute the final test error and plot the learned function against the data and the true underlying function
2. Repeat for `N = 200, 500, 1000 and 2000` (I suggest you program a function with N as a parameter)
3. Compare the results:
      a) Are there differences in the models for different `N`? (size, decay)
      b) Are there differences in prediction performance for different `N`? In other words, do the models get better, worse, or indifferent?

***

### Exercise 3

This exercise involves the use of the South African Coronary Heart Disease (CHD) data set, which we have seen in previous labs.
The task is to predict CHD using age at onset, current alcohol consumption, obesity levels, cumulative tobacco, type-A behavior, and low density lipoprotein cholesterol as predictors

```{r}
set.seed (3)

SAheart <- read.csv("MDSAheart.data")

## First some necessary pre-processing (this part should be lenghtier as you know but let's go to the point)

SAheart <- subset(SAheart, select=-row.names)
SAheart[,"famhist"] <- as.factor(SAheart[,"famhist"])
SAheart[,"chd"] <- factor(SAheart[,"chd"], labels=c("No","Yes"))

## There are no missing values; alcohol, ldl and tobacco have very skewed distributions, so we log transform them

SAheart[,"tobacco"] <- log(SAheart[,"tobacco"]+0.01)
SAheart[,"alcohol"] <- log(SAheart[,"alcohol"]+0.01)
SAheart[,"ldl"] <- log(SAheart[,"ldl"])

## We now standardize all continuous predictors

SAheart[,-c(5,10)] <- scale(subset(SAheart, select=-c(5,10)))
```


So this is your departing data set

```{r}
summary(SAheart) 
```


Do the following (in separate sections):

1. Create learning and test sets, selecting randomly 2/3 and 1/3 of the data
2. Fit a MLP model using only the learn data with Coronary Heart Disease (chd) as the target
   using 10x10 CV
3. Calculate the misclassification rate for your model in the learn data and in the leftout data (test data)
   If you finished the previous assignment with this dataset (L3), you can now compare the obtained results

Note this is a very challenging data set

Your code starts here ...

