---
title: 'LAB 4: Linear regression and beyond'
subtitle: 'Exercise: Analysis of daily air quality measurements in New York'
author: "Lluís A. Belanche, translated to R Markdown by Marta Arias"
date: "March 2020"
output:
  html_notebook: default
pdf_document: default
editor_options: 
  chunk_output_type: inline
---


```{r}
summary(airquality)
```

For information on the dataset
```{r}
?airquality
```

The goal is to develop a predictive model of Ozone concentration given solar radiation, wind and temperature

Do the following (in separate sections):

1. Fix the missing values; most are located in the target variable itself (Ozone); since this is a time series, you can either remove the rows or replace the missing values by interpolation; then deal with Solar.R (knn)

2. Do any other pre-processing or visual inspection of the dataset that you think is adequate. Split the data into training (2/3) and test (1/3); WARNING: do this randomly because the dataset has an structure (days are consecutive, etc) or shuffle the data before partitioning

3. Fit a `lm()` model to predict 'Ozone' with solar radiation, wind and temperature in the training set

4. Inspect the results and compute the predictive normalized MSE

5. Try a second model by log-transforming the Ozone; is this an improvement? if so, keep it

6. Try a third model that additionally includes 'Day'; is this an improvement? if so, keep it

7. Try a last model that includes a second order polynomial on Wind, like this:
`lm(log(Ozone) ~ Solar.R + Temp + poly(Wind,2), data = airquality)`
is this an improvement? if so, keep it

8. Do another modelling using ridge regression, and decide the best lambda using GCV

9. Do another modelling using LASSO regression

10. Decide which is your best model and give a final prediction in the test set

#### Your code starts here ...
